An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2011-03-16 paper.tex
Partially complete edit, pushing for Ilya.

\documentclass[prd,preprint]{revtex4}

@@

\bibliographystyle{h-physrev}

@@

\begin{abstract}
  Selection among alternative theoretical models given an observed
  data set is an important challenge in many areas of physics and
  astronomy, ranging from gravitational-wave data analysis to
  measurements of black-hole masses in X-ray binaries.
  Reversible-jump Markov chain Monte Carlo (RJMCMC) is an extremely
  powerful technique for performing Bayesian model selection, but it
  suffers from a fundamental difficulty: it requires jumps between
  model parameter spaces, but cannot retain a memory of the favored
  locations in more than one parameter space at a time.  Thus, a naive
  jump between parameter spaces is unlikely to be accepted in the MCMC
  algorithm and convergence is correspondingly slow.  Here we
  demonstrate an interpolation technique that uses samples from
  single-model MCMCs to propose inter-model jumps from an
  approximation to the single-model posterior of the target parameter
  space.  The interpolation technique, based on a kD-tree data
  structure, is adaptive and efficient in arbitrary dimensions.  We
  show that our technique leads to dramatically improved convergence
  over naive jumps in an RJMCMC, and compare it to other proposals in
  the literature to improve the convergence of RJMCMCs.  We also
  discuss the use of the same interpolation technique in two other
  contexts: as a convergence test for a single-model MCMC and as a way
  to construct efficient ``global'' proposal distributions for
  single-model MCMCs without prior knowledge of the structure of the
  posterior distribution.
\end{abstract}

@@

\will{Mention Weinberg direct integration (Ref.~\cite{Weinberg2009}).
  Also \cite{Ascasibar2005} in the cosmological context (not really
  kD-tree, but close).}

\section{Reversible Jump MCMC}

@@

A Markov chain Monte Carlo \cite{Gilks:1996} produces a set of samples $\{ \vtheta^{(j)} \, | \, i = 1, \ldots \}$ from the model parameter space that are sampled according to the posterior, meaning that, in the limit that the chain length tends to infinity, the relative frequency with which a given set of parameter appears in the chain is proportional to the desired posterior, $p(\vtheta|d,M)$.  Therefore, the output of an MCMC can be directly interpreted as the posterior PDF over the full parameter space, while PDFs for individual parameters can be obtained through trivially marginalizing over the uninteresting parameters, simply by considering the distribution of the parameter of interest in the MCMC samples.

A Markov chain has the property that the probability distribution of the next state can depend only on the current state, not on the past history:
\be
p(\vtheta^{(j+1)})=\int_{V} d\vtheta^{(j)} p(\vtheta^{(j)} \to \vtheta^{(j+1)}) p(\vtheta^{(j)}),
\ee
where the jump probability $p(\vtheta^{(j)} \to \vtheta^{(j+1)})$ depends only on $\vtheta^{(j)}$ and $\vtheta^{(j+1)}$. 
An additional requirement for an MCMC arises from the fact that the desired distribution is the equilibrium distribution.  In other words, if we assume that state $(j)$ of the chain is sampled from the desired PDF, $p(\vtheta^{(j)})=p(\vtheta|d,M)$ , then the next state $(j+1)$ must be sampled from the PDF as well, so that $p(\vtheta^{(j+1)})=p(\vtheta|d,M)$.

One way to produce such a sequence of samples is via the Metropolis-Hastings algorithm, first proposed by Metropolis et al.~in 1953 \cite{Metropolis:1953}, and later generalized by Hastings:
\begin{enumerate}
  \item Given a current state $\vtheta^{(j)}$, propose the next state $\vtheta^p$ by drawing from a jump proposal distribution with probability $Q(\vtheta^{(j)} \to \vtheta^p)$.  
  \item Compute the probability of accepting the proposed jump as
\bel{eq:p-accept}
p_{\textnormal{accept}}  \equiv \min\Bigl(1,  \frac{p(\vtheta^p|d)}{p(\vtheta^{(j)}|d)} \frac{Q(\vtheta^p \to
        \vtheta^{(j)})}{Q(\vtheta^{(j)} \to \vtheta^p)} \Bigr).
\ee
\item Pick a uniform random number $\alpha \in [0,1]$.  If $\alpha<  p_{\textnormal{accept}}$, accept the proposed jump, setting $\vtheta^{(j+1)}=\vtheta^p$.  Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^{(j+1)}=\vtheta^{(j)}$.
 \end{enumerate}

This jump proposal distribution $Q(\vtheta^{(j)} \to \vtheta^p)$ can depend on the parameters of the current state $\vtheta^{(j)}$, but not on the past history.  It must also allow any state within the allowed prior volume to be reachable by the MCMC.  Beyond these conditions, there is obviously a great deal of freedom in choosing the jump proposal distribution.  This is the most important choice in the MCMC, as it determines the sampling efficiency of the algorithm, i.e., the necessary length of the chain before it converges to the posterior PDF.  Creating an efficient jump proposal distribution requires an understanding of the structure of the parameter space which may not be available until the PDFs are found, creating a Catch-22; one possibility for resolving this infinite loop is described in Section \ref{sec:examples}.

@@

Finally, we note that, in practice, the randomly chosen initial
starting point of the MCMC may be in a particularly unlikely location
in the parameter space.  Because jumps are frequently local, we will
generally want to ignore the first few autocorrelation lengths worth
of points in a finite-size chain to avoid biases in the recovered
posterior PDF due to the choice of the initial location.  The points
thus discarded are referred to as "burn-in" points.

@@

The samples produced during by an MCMC algorithm can be used to
directly perform a Monte Carlo evidence integral.  This results in a
harmonic mean estimator for the evidence, which suffers from a large
variance and a possible bias
\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  Additional
techniques for the direct integration of evidence, also based on a kD
tree decomposition of the parameter space (see Sec.~\ref{sec:kDTree}),
are described in \cite{Weinberg2009}.  These techniques are promising,
but in some cases suffer \cite{Farr2010} from large variance and bias
as well.  An alternative approach to model selection among a set of
models is based on performing an MCMC in a ``super-model'' that
encompasses all of the models under consideration; this is known the
the Reversible Jump Markov chain Monte Carlo (RJMCMC).

The parameter space of the super-model in an RJMCMC consists of a
discrete parameter that identifies the model, $M_i$, and a set of
continuous parameters appropriate for that model, $\vtheta_i$.  Thus,
each sample consists of a model identifier and a location within the
parameter space of that model, $\{M_i, \vtheta_i\}$.  We perform the
MCMC in the ``super-model'' parameter space just like a regular MCMC;
in general, we can both propose jumps to different parameters within a
model (intramodel jumps) and jumps to a different model with different
parameters (intermodel jumps).  The resulting chain samples from the
posterior $p(M_i,\vtheta_i|d)$.  As in a usual MCMC, the PDF on the
model as a parameter, with other parameters ignored, is obtained by
marginalizing over the remaining parameters, i.e., the evidence for a
given model is just
%
\be
p(M_i|d) = \int d\vtheta_i p(M_i, \vtheta_i|d) \approx \frac{N_i}{N},
\ee
%
where $N_i$ is the number of RJMCMC samples listing the $i$'th model
and $N$ is the total chain length.  Thus, the probability of a
particular model relative to other models under consideration is given
by the fraction of RJMCMC samples lying in the parameter space of that
model.

@@

\section{RJMCMC Efficiency}

@@

\section{Conclusion/Discussion}

\nocite{Littenberg2009}

>>>>

\documentclass[aps,prd,reprint]{revtex4-1}

@@

\bibliographystyle{apsrev4-1}

@@

\begin{abstract}
  Selection among alternative theoretical models given an observed
  data set is an important challenge in many areas of physics and
  astronomy.
  %ranging from gravitational-wave data analysis to
  %measurements of black-hole masses in X-ray binaries.
  Reversible-jump Markov chain Monte Carlo (RJMCMC) is an extremely
  powerful technique for performing Bayesian model selection, but it
  suffers from a fundamental difficulty: it requires jumps between
  model parameter spaces, but cannot retain a memory of the favored
  locations in more than one parameter space at a time.  Thus, a naive
  jump between parameter spaces is unlikely to be accepted in the MCMC
  algorithm and convergence is correspondingly slow.  Here we
  demonstrate an interpolation technique that uses samples from
  single-model MCMCs to propose inter-model jumps from an
  approximation to the single-model posterior of the target parameter
  space.  The interpolation technique, based on a kD-tree data
  structure, is adaptive and efficient in arbitrary dimensions.  We
  show that our technique leads to dramatically improved convergence
  over naive jumps in an RJMCMC, and compare it to other proposals in
  the literature to improve the convergence of RJMCMCs.  We also
  discuss the use of the same interpolation technique in two other
  contexts: as a convergence test for a single-model MCMC and as a way
  to construct efficient ``global'' proposal distributions for
  single-model MCMCs without prior knowledge of the structure of the
  posterior distribution.
\end{abstract}

@@

Selection among alternative theoretical models given an observed data
set is an important challenge in many areas of physics and astronomy.
In a Bayesian context, model selection involves computing the evidence
for each model given the data.  The model evidence is an integral of
the unnormalized posterior probability distribution over the model
parameter space, representing the ``total probability'' in that model.
Models with larger evidence are preferred; the ratio of the evidences
of two models is the Bayes ratio between them.  The product of the
Bayes ratio and the ratio of prior probabilities for the model gives
the odds ratio for the models, the ratio of Bayesian probabilities for
the two models given the observation data.

There are many ways to compute model evidences.  In low-dimensional
parameter spaces, the unnormalized posterior probability can be
evaluated on a grid or lattice and the integral can be performed
directly.  For many problems or models of interest, however, the
dimensionality of parameter space is too large to make this approach
practical.  Nested sampling \cite{Skilling:2006,Feroz:2009} \ilya{More
  references?} is a stochastic method for exploring the parameter
space of a model while computing the evidence integral statistically
\ilya{I want a better word than ``statistically'' to indicate that
  it's doing something akin to Monte-Carlo integration (but I would
  like to reserve Monte-Carlo for MCMC).}.

\ilya{Something about how nested sampling distributes points according
  to the prior, not the posterior, and therefore can miss modes that
  are disfavored by prior.  Also, it does not exploit any information
  about the posterior to make the evidence calculation more
  efficient.}

In contrast to nested sampling, where computation of evidences is the
primary objective, Markov Chain Monte Carlo (MCMC) methods attempt to
stochastically produce parameter samples with density proportional to
the posterior probability distribution.  In MCMC techniques, the
primary target is an accurate estimate of the posterior distribution.
It is not straightforward to compute the model evidence from the MCMC
samples.  The most direct way to estimate the evidence for a model
from MCMC samples is to compute the harmonic-mean estimatator, but
this estimator of the evidence suffers from infinite
variance\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}; various
methods have been proposed to stabilize it \will{Citation}, but all
require some assumptions about the shape of the posterior distribution
that may not be satisfied in practice.

Reference \cite{Weinberg2009} gives a method for directly computing
the evidence integral using a kD-tree datastructure to decompose a
parameter space into boxes containing the MCMC sample points.  The
integral is approximated as a sum over box volumes.  This method is
promising, but it is not clear in general what statistical and
systematic errors it introduces and how these are affected by the
shape of the posterior distribution from which the MCMC samples.

When the goal is model selection between several known models, only
the \emph{relative} evidence of each model is needed.  In this
circumstance, the Reversible Jump MCMC technique first introduced in
Reference \cite{Green1995} is one of the most reliable and accurate
ways to compare the models.  Reversible Jump MCMC (RJMCMC), described
more fully in Section \ref{sec:reversible-jump}, performs a standard
MCMC in a superspace that is a direct sum of all the model parameter
spaces.  Such an MCMC involves both intra- and inter-model jumps; the
number of MCMC samples in each model's parameter space is proportional
to that model's relative evidence in the suite of models being
compared.

Implemented naively, RJMCMC has a significant drawback: because the
chain of samples must be Markovian, only the current sample is
available to the algorithm.  Each time an RJMCMC transitions between
models, the information about the good parameter values in the
previous model is lost; subsequent jumps into that model must ``start
fresh,'' and are correspondingly unlikely to be accepted, delaying
convergence of the RJMCMC sample chain.

Here we demonstrate a technique for improving the acceptance ratio of
inter-model jumps in an RJMCMC, leading to dramatically improved
convergence of RJMCMC sample chains.  The technique uses a kD-tree
datastructure to construct an approximation to each model's posterior
parameter distribution.  We propose jumps into each model drawn from
the approximation to the posterior.  Because jumps are proposed
preferentially to locations favored by the single-model posterior, the
RJMCMC compares ``good'' locations in parameter space across all the
models, and convergence is generally rapid.  The approximation to the
posterior distribution produced by the kD-tree is a constant-in-box
interpolation of the posterior, similar in spirit to the phase-space
density interpolants produced from N-body positions and momenta in
Refernce \cite{Ascasibar2005}.  The kD-tree interpolation is effective
in arbitrary numbers of dimension, and is quite space-efficient,
requiring $\order{N \log N}$ storage space and $\order{\log N}$ time
to produce each proposed jump, where $N$ is the number of single-model
MCMC samples used to construct the interpolation.  Reference
\cite{Littenberg2009} proposed an alternate method for producing
inter-model jumps in an RJMCMC that relies on a box decomposition of
parameter space, but uses fixed-sized boxes; because the box size is
fixed, the method cannot adapt to the local structure of the
posterior, and therefore becomes asymptotically less efficient than
the adaptive kD-tree decomposition in either acceptance ratio or
storage space as the parameter-space dimension increases or the
posterior becomes more peaked.  \ilya{Mention something (with
  reference) about the 10-way RJMCMC in the black-hole mass paper?}

The structure of this paper is as follows: in Section
\ref{sec:reversible-jump} we introduce in more detail the concept of a
Reversible Jump MCMC, and describe the fundamental difficulty with a
naive jump proposal in an RJMCMC.  In Section \ref{sec:kDTree} we
introduce the kD-tree datastructure used to decompose parameter space
into boxes for the interpolation.  In Section \ref{sec:efficiency} we
demonstrate the efficiency gains that are achieved from use of the
interpolated jump proposal.  In Section \ref{sec:examples} we give
examples of some other uses of the interpolated jump proposal in a
single-model MCMC context.  Finally, in Section \ref{sec:conclusion}
we offer a summary and some concluding remarks on the method.

\section{Reversible Jump MCMC}
\label{sec:reversible-jump}

@@

A Markov chain Monte Carlo \cite{Gilks:1996} produces a set of samples
$\{ \vtheta^{(j)} \, | \, i = 1, \ldots \}$ from the model parameter
space that are sampled according to the posterior, meaning that, in
the limit that the chain length tends to infinity, the relative
frequency with which a given set of parameter appears in the chain is
proportional to the desired posterior, $p(\vtheta|d,M)$.  Therefore,
the output of an MCMC can be directly interpreted as the posterior PDF
over the full parameter space, while PDFs for individual parameters
can be obtained by marginalizing over the uninteresting parameters.

A Markov chain has the property that the probability distribution of
the next state can depend only on the current state, not on the past
history: 
%
\be 
p(\vtheta^{(j+1)})=\int_{V} d\vtheta^{(j)} p(\vtheta^{(j)} \to
\vtheta^{(j+1)}) p(\vtheta^{(j)}), \ee
%
where the jump probability $p(\vtheta^{(j)} \to \vtheta^{(j+1)})$
depends only on $\vtheta^{(j)}$ and $\vtheta^{(j+1)}$.  An additional
requirement for an MCMC arises from the fact that the desired
distribution is the equilibrium distribution.  In other words, if we
assume that state $(j)$ of the chain is sampled from the desired PDF,
$p(\vtheta^{(j)})=p(\vtheta|d,M)$ , then the next state $(j+1)$ must
be sampled from the PDF as well, so that
$p(\vtheta^{(j+1)})=p(\vtheta|d,M)$; this condition is known as
``detailed balance.''

One way to produce such a sequence of samples is via the
Metropolis-Hastings algorithm, first proposed in Reference
\cite{Metropolis:1953}, and later generalized by Hastings
\ilya{Citation?}:
\begin{enumerate}
\item Given a current state $\vtheta^{(j)}$, propose the next state
  $\vtheta^p$ by drawing from a jump proposal distribution with
  probability $Q(\vtheta^{(j)} \to \vtheta^p)$.
\item Compute the probability of accepting the proposed jump as
  \bel{eq:p-accept} 
p_{\textnormal{accept}} \equiv \min\Bigl(1,
\frac{p(\vtheta^p|d)}{p(\vtheta^{(j)}|d)} \frac{Q(\vtheta^p \to
  \vtheta^{(j)})}{Q(\vtheta^{(j)} \to \vtheta^p)} \Bigr).  
\ee
\item Pick a uniform random number $\alpha \in [0,1]$.  If $\alpha<
  p_{\textnormal{accept}}$, accept the proposed jump, setting
  $\vtheta^{(j+1)}=\vtheta^p$.  Otherwise, reject the jump, and remain
  at the same location in parameter space for the next step,
  $\vtheta^{(j+1)}=\vtheta^{(j)}$.
\end{enumerate}

This jump proposal distribution $Q(\vtheta^{(j)} \to \vtheta^p)$ can
depend on the parameters of the current state $\vtheta^{(j)}$, but not
on the past history.  It must also allow any state within the allowed
prior volume to be reachable by the MCMC.  Any jump proposal that
satisfies these properties is suitable for an MCMC.

The jump proposal is the most important choice in the MCMC, as it
determines the sampling efficiency of the algorithm, i.e., the
necessary length of the chain before it converges to the posterior
PDF.  Creating an efficient jump proposal distribution requires an
understanding of the structure of the parameter space which may not be
available until the PDFs are found, creating a Catch-22; one
possibility for resolving this infinite loop is described in Section
\ref{sec:examples}.

@@

Finally, we note that, in practice, the randomly chosen initial
starting point of the MCMC may be in a particularly unlikely location
in the parameter space.  Because jumps are frequently local, we will
generally want to ignore the early points in a finite-size chain to
avoid biases in the recovered posterior PDF due to the choice of the
initial location.  The points thus discarded are referred to as
"burn-in" points.

@@

The samples produced by an MCMC algorithm can be used to directly
perform a Monte Carlo evidence integral.  This results in a harmonic
mean estimator for the evidence, which suffers from a large variance
and bias \cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.
Additional techniques for the direct integration of evidence, also
based on a kD tree decomposition of the parameter space (see
Sec.~\ref{sec:kDTree}), are described in \cite{Weinberg2009}.  These
techniques are promising, but in some cases suffer \cite{Farr2010}
from large variance and bias as well.  An alternative approach to
model selection among a set of models is based on performing an MCMC
in a ``super-model'' that encompasses all of the models under
consideration; this is known the the Reversible Jump Markov chain
Monte Carlo (RJMCMC).

The parameter space of the super-model in an RJMCMC consists of a
discrete parameter that identifies the model, $M_i$, and a set of
continuous parameters appropriate for that model, $\vtheta_i$.  Thus,
each sample consists of a model identifier and a location within the
parameter space of that model, $\{M_i, \vtheta_i\}$.  We perform the
MCMC in the ``super-model'' parameter space just like a regular MCMC;
in general, we can both propose jumps to different parameters within a
model (intramodel jumps) and jumps to a different model with different
parameters (intermodel jumps).  The resulting chain samples from the
posterior $p(M_i,\vtheta_i|d)$.  As in a usual MCMC, the PDF on the
model as a parameter, with other parameters ignored, is obtained by
marginalizing over the remaining parameters.  If we assume uniform
priors on the models, the evidence for a given model is proportional
to the number of counts 
%
\be
p(M_i|d) = \int d\vtheta_i p(M_i, \vtheta_i|d) \approx \frac{N_i}{N},
\ee
%
where $N_i$ is the number of RJMCMC samples listing the $i$'th model
and $N$ is the total chain length.  Thus, the probability of a
particular model relative to other models under consideration is given
by the fraction of RJMCMC samples lying in the parameter space of that
model.

@@

\section{RJMCMC Efficiency}
\label{sec:efficiency}

@@

\section{Conclusion/Discussion}
\label{sec:conclusion}
